Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-03-31 talk/icooolps2009/paper.tex
accepted most of anto's changes

\anto{XXX: are the following paragraphs really needed? I don't think this
  details are needed in Section \ref{sect:implementation}.  I would say
  something much shorter, see below.}

The central idea of this way to implement VMs is that the interpreter
implementation in RPython should be as free as possible of low-level
implementation details, such as memory management strategy, threading model or
object layout. Instead, these details are inserted into the VM during the
translation process by the translation toolchain. This makes it possible to
change these details later, if that becomes necessary. This is something that is
hard to do with a traditional VM written in a low-level language such as C,
since the low-level details need to be fixed early in the development-process.
XXX is this paragraph really needed?

In the following we will describe some details of the translation process, since
they are needed in Section \ref{sect:implementation}. The first step is to produce
control flow graphs of all functions of the RPython program. Afterwards, type
inference is performed to gain type information about all the variables in the
flow graphs. Afterwards, the abstraction level of the operations in the graphs
is lowered in a stepwise fashion. At the end of this process, all operations in
the graphs correspond rather directly to a simple operation in C. The variables
are annotated with a C-like type system containing primitive types (like
\texttt{Signed}, \texttt{Bool}, etc.) or pointer types pointing to Structs,
Arrays or Functions.

XXX example. reuse the one of the tracing jit?

The translation process usually just turns these graphs into C code so that they
can be compiled into an executable. However, they can also be interpreted in
various ways. This is useful for testing and debugging the translation toolchain
because the produced error messages in case of a crash are a lot more helpful
than what would be produced after compilation to C. These low-level graphs are
also what the tracing JIT takes as input, as we will see later.

\anto{By writing VMs in a high-level language, we keep the implementation of
  the language free of low-level details such as memory management strategy,
  threading model or object layout.  These features are automatically added
  during the translation process which consists in a series of steps, each
  step transforming the representation of the program produced by the previous
  one until we get the final executable.  As we will see later, this internal
  low-level representation of the program is also used as an input for the
  tracing JIT.}

@@

\sout{The generation of loops works as follows: at first, everything is interpreted.}
\anto{At first, when the program starts, everything is interpreted.}
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

When a \sout{common}\anto{hot} loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

>>>>

By writing VMs in a high-level language, we keep the implementation of the
language free of low-level details such as memory management strategy,
threading model or object layout.  These features are automatically added
during the translation process which consists in a series of steps, each step
transforming the representation of the program produced by the previous one
until we get the final executable.  As we will see later, this internal
low-level representation of the program is also used as an input for the
tracing JIT.

@@

At first, when the program starts, everything is interpreted.
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.
